Dependent Types and Martin-L\"of Type Theory: An Introduction

### 2016-07-03 paper.tex
yo

% Some characters that are not automatically defined
% (you figure out by the latex compilation errors you get),
% and you need to define:
\DeclareUnicodeCharacter{8988}{\ensuremath{\ulcorner}}
\DeclareUnicodeCharacter{8989}{\ensuremath{\urcorner}}
\DeclareUnicodeCharacter{8803}{\ensuremath{\overline{\equiv}}}
\DeclareUnicodeCharacter{120738}{\ensuremath{\Sigma}}
\DeclareUnicodeCharacter{955}{\ensuremath{\lambda}}

@@

\section{Martin-L\"of Type Theory}
%% Curry-Howard?
In this section, we introduce the foundational concepts of Martin-L\"of Type
Theory. We discuss recursion, induction, and as well several primitive type
encodings (e.g. from propositional logic). We recommend the reader to have a
running proof assistant while reading this section to help fortify (better word)
several concepts.

\subsection*{Principles}
Recursion and induction are one of the powerful tools available in a
fully-dependently typed system. Using recursion and induction principles to
define functions and proofs, we are guaranteed termination and totality. This is
because, in general, recursion and induction principles are defined under certain
known intuitions on the given type that they are meant to act on and thus also
follow a certain pattern.
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
recT : (C : Set) → C₀ ... → T → C
indT : (C : T → Set) → (C₀ : ...) ... → (t : T) → C t
\end{code}
\end{minipage}

\begin{minipage}{0.7\textwidth}
\begin{small}

The recursion and induction principles for some type {\tt T}.
\end{small}
\end{minipage}
\end{center}
The first argument given to these principles is sometimes called the {\em motive}.
This simply means that {\tt C} is the type we are going to inhabit. For recursion
principles, this means the given output of the function we are defining, and for
induction principles, this can be read as the proposition we are trying to prove.
The terms that follow (viz. {\tt C₀}) correspond to the constructors of the given
type, which is how these principles guarantee totality. Finally, both principles
take a term of the type that are acting on (viz. {\tt T} and {\tt (t : T)}). The
only difference is that the induction principle introduces a type family, as
opposed to non-parameterized type in the recursion principle. We discuss this in
Section 4.

@@

\subsection*{Primitive Type Encodings}
%% Bottom, Top, Disjunction, Conjunction, Sigma

@@

\subsection*{Booleans}
\subsection*{Natural Numbers}
\subsection*{List/Orderings}
\subsection*{Identity Types}

>>>>

% Some characters that are not automatically defined
% (you figure out by the latex compilation errors you get),
% and you need to define:
\DeclareUnicodeCharacter{8988}{\ensuremath{\ulcorner}}
\DeclareUnicodeCharacter{8989}{\ensuremath{\urcorner}}
\DeclareUnicodeCharacter{8803}{\ensuremath{\overline{\equiv}}}
\DeclareUnicodeCharacter{120738}{\ensuremath{\Sigma}}
\DeclareUnicodeCharacter{955}{\ensuremath{\lambda}}
\DeclareUnicodeCharacter{2261}{\ensuremath{\equiv}}

@@

\section{Martin-L\"of Type Theory}
%% Curry-Howard?
In this section, we introduce the foundational concepts of Martin-L\"of Type
Theory \cite{}. We discuss recursion, induction, and as well several primitive type
encodings (e.g. from propositional logic).

\subsection*{Principles}
Recursion and induction are one of the powerful tools available in a
fully-dependently typed system. Using recursion and induction principles to
define functions and proofs, we are guaranteed termination and totality. This is
because, in general, recursion and induction principles are defined under certain
known intuitions (we can think of them as axioms) on the given type that they are
meant to act on. They also follow a certain pattern:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
recT : (C : Set) → C₀ ... → T → C
indT : (C : T → Set) → (C₀ : ...) ... → (t : T) → C t
\end{code}
\end{minipage}

\begin{minipage}{0.7\textwidth}
\begin{small}
The recursion and induction principles for some type {\tt T}.
\end{small}
\end{minipage}
\end{center}
The first argument given to these principles is sometimes called the {\em motive}.
This simply means that {\tt C} is the type we are going to inhabit. For recursion
principles, this means the given output of the function we are defining, and for
induction principles, this can be read as the proposition we are trying to prove.
The terms that follow (viz. {\tt C₀}) correspond to the constructors of the given
type, which is how these principles guarantee totality. Finally, both principles
take a term of the type that are acting on (viz. {\tt T} and {\tt (t : T)}). The
only difference is that the induction principle introduces a type family, as
opposed to non-parameterized type in the recursion principle. We discuss this in
Section 4.

@@

\subsection*{Primitive Type Encodings}
%% Bottom, Top, Disjunction, Conjunction, Sigma
We start with \(\perp\), the empty type. We include the definition of \(\perp\),
{\tt rec⊥} and {\tt ind⊥}:

@@

\subsection*{Booleans}
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
data 𝔹 : Set where
  True  : 𝔹
  False : 𝔹

rec𝔹 : (C : Set) → C → C → 𝔹 → C
rec𝔹 C t f True = t
rec𝔹 C t f False = f

ind𝔹 : (C : 𝔹 → Set) → C True → C False → (x : 𝔹) → C x
ind𝔹 C t f True = t
ind𝔹 C t f False = f
\end{code}
\end{minipage}
\end{center}

\subsection*{Natural Numbers}
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
recℕ : (C : Set) → C → (ℕ → C → C) → ℕ → C
recℕ C c _ 0 = c
recℕ C c f (suc n) = f n (recℕ C c f n)

indℕ : (C : ℕ → Set) → C 0 →
       (∀ (x : ℕ) → C x → C (suc x)) →
       (x : ℕ) → C x
indℕ C c _ 0 = c
indℕ C c f (suc n) = f n (indℕ C c f n)
\end{code}
\end{minipage}
\end{center}

\subsection*{List/Orderings}
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
data List (A : Set) : Set where
  Nil  : List A
  _::_ : A → List A → List A

recList : {A : Set} → (C : Set) → C →
          (A → List A → C → C) → List A → C
recList C c f Nil = c
recList C c f (x :: xs) = f x xs (recList C c f xs)

indList : {A : Set} → (C : List A → Set) → C Nil →
          ((x : A) → (xs : List A) → C xs → C (x :: xs)) →
          (xs : List A) → C xs
indList C c f Nil = c
indList C c f (x :: xs) = f x xs (indList C c f xs)
\end{code}
\end{minipage}
\end{center}

\subsection*{Identity Types}
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
infix  4  _≡_     -- propositional equality

data _≡_ {ℓ} {A : Set ℓ} : (a b : A) → Set ℓ where
  refl : (a : A) → (a ≡ a)

ind≡ : {A : Set} (C : {x y : A} → x ≡ y → Set) →
       (∀ (x : A) → C (refl x)) →
       {x y : A} (p : x ≡ y) → C p
ind≡ C c (refl x) = c x

rec≡ : {A : Set} {x y : A} (C : A → Set) → (p : x ≡ y) → C x → C y
rec≡ {A} {x} {y} C = 
  ind≡ (λ {x} {y} _ → C x → C y)
       (λ x → λ z → z)
       {x} {y} 
\end{code}
\end{minipage}
\end{center}
